Was Devon Allen Unjustly Disqualified at the 2022 World Track and Field Championships?

### 2023-04-24 Manuscript/manuscript.tex
just started reading intro, have not finished

\title{On Devon Allen's Disqualification at the 2022 Track and Field World
Championship}

@@

%Abstract should be 200 words
%Devon Allen was disqualified in 2022... discussions about dq, talk methods and 
%models, venue effect, and conclusion
\begin{abstract}
  Devon Allen was disqualified in the Men's 110 meter hurdle final of the 2022
  Track and Field World Championships after registering a reaction time of 0.099 seconds, 0.001 
  seconds faster than what is allowed.  Following the games, bloggers on the running website 
  LetsRun scrutinized the results and performed non-statistical analysis of the data.
  They found that the reaction time data from the 2022 World Track Championships seemed to be faster 
  compared to the other datasets they looked at, but did not perform any meaningful statistical 
  analysis of the data.  This paper questions the reaction time disqualification barrier, which is
  currently 0.1 seconds to determine whether this is a reasonable threshold.
  We employ a generalized linear mixed model (GLMM) with a random effect term
  which will be referred to as the venue effect in order to try and model the
  data.  Additionally, we employ a signed-rank test for clustered data to compare
  reaction times for the same athletes at different competitions.
  This matter needs to be addressed because Devon Allen's disqualification could 
  be a problem that could continue to re-appear in future world championships.

\bigskip
\noindent{\sc Keywords}:
Reaction Time, Seiko,

@@

\jy{The bib file needs quality control. Practice my writing tips in the
  stat-wriging notes: chapter 2 on latex/bibtex.}

\jy{Go over the reference section to see what need to be changed in the bib source}

@@

After coming in third in the 110 meter hurdles at the United States Track and 
Field Championships in Eugene Oregon at the end of June, Devin Allen was expected to
be a contender for the World Championships, which were scheduled for two months 
later, also in Eugene. At the World Championships, Allen made it through the 
heats and semifinals to reach the finals while posting reaction times of 0.123
in the heats and 0.101 in the semifinal.  Allen, who attended the University
of Oregon (Location of US World Championship and the World Championship) was
considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
in early June; only 0.04 seconds away from the world record \citep{wa2022preview}.
Thus his disqualification by reacting in 0.099 seconds instead of 0.01 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that show the disqualification and can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers when
they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting another two years until the next World Track and Field Championship.
An investigation into reaction times is also important because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times; which is the goal of
every sprinter who steps onto the track.

Allen's disqualification has been discussed extensively in online communities
such as \url{www.LetsRun.com}, a website that is part message board and part
news. The message board,
which functions similarly to Reddit is very active during important running
events, which include during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
\jy{cite these blog articles formally}
news section of the website that generated articles such as ``Was Devon Allen
Screwed? There's At Least a 99\% That He Was" and ``The Data Keeps Pouring In and
It Continues to Look Bad For World Athletics and Great For Devon Allen".  Robert
Johnson, who created and runs the Letsrun website argued that there was an error
with the timing equipment at the 2022 Track and Field World Championship and
cited graphs, statistics such as median reaction time, and compared the reaction
times at the 2022 Track and Field World Championship \citep{johnson2022devon}.

Reaction time in sprint and hurdle events has been studied by many authors.
\citet{pain2007sprint} looked at
reaction times for nine male athletes and analyzed their reaction times.  The
fastest athlete of the nine, had on average (even with their two fastest time 
removed) a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions.  In a study commissioned by IAAF,
\jy{define acronym at first occurrence}
\citep{komi2009iaaf} recommended that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body
movement. This could provide a more accurate representation of 
reaction time, and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been proved to be a reliable
measure of reaction time \citep{mudric2015evaluation}.

\jy{Could be condensed}
\citet{pilianidis2012start} found
significant differences between reaction times at several world championships
between 1997 and 2009 for the men's 110 meter hurdles.  Another important
conclusion was that for reaction times did not decrease over the twelve years
covered by the research in the study \citep{pilianidis2012start}.  This contradicts
what was found in a 2021 study that concluded the year of competition has a
significant effect on reaction times \citep{zhang2021correlation}.  Their
analysis looked at data from the 2011-2019 World Championships across multiple
sprint events and concluded that from 2011 to 2019, reaction times got
faster. However, not  all studies have been in favor of a decrease in the
reaction time barrier. \citet*{brosnan2017effects} in 2016 argue for increasing 
the reaction time barrier to 0.115 seconds, and cited evidence from European 
Championships and World Championships in their article.  A study looking 
at the 2008 Olympics likewise supported the conclusion that athletes could not
react in as little as 0.1 seconds \citep{lipps2011implications}.  Another study
concluded that visual reaction times to the starting gun may be faster by 0.007
seconds than the IAAF method, which is from force on block starts 
\citep{holmes2018method}.  While this may seem trivial, Allen was disqualified
only by 0.01 seconds. It is also worth noting that one study looking at data 
from various IAAF meets found that the fastest reaction times for men were found
between the ages of 26 and 29 \citep{tonnessen2013reaction}.  This is notable 
because Allen was 27 years old at the 2022 Track and Field World Championship 
and in his physical peak.

\jy{language needs to be formalized and condensed.}
The objective to this paper is to analyze the reaction times at the 2022
World Track and Field Championship to check for any abnormalities, and we approach
this two different ways.  Our first method is to try to model the data, which is
the reaction times of every non-disqualification since the 1999 World Championships. 
The model we use is a gamma linear mixed-effects model with a random effect
accounting for the venue (or year) effect. 
The second method is to compare data for athletes who have competed
at additional competitions to see how their reaction times compare from previous
competitions to this one. First we can look at the reaction times of athletes who
competed at the World Track and Field Championship in both 2019 and 2022 and
compare the times, and secondly we can look at how athletes performed compared
to how they did at their national/qualifiers meet. In order to reduce any chances
of athletes improving reaction time, we can also look at data from qualifying meets.
Almost every athlete had to qualify by proving they are elite at the national level, 
usually by coming in at least fourth. Thus, data was compiled on athletes who competed
at the 2022 World Track and Field Championships on what these athletes also ran 
at their country's national championship. If there was any argument that athletes
have improved their reaction time from 2019 to 2022, it should now be considered
nearly null, as the national competitions took place between June and July 2022 
and the World Championships were in August. The equipment used to measure reaction
time cannot be guaranteed to be the same from national to international
competition, but the time span between meets has been reduced considerably.

The rest of this paper is as follows: Section~\ref{sec:Data} describes how data was
collected and begins to detail the generalized linear mixed-effect (GLMM)
model which is further developed in the
Section~\ref{sec:Methods}.  The signed-rank test analysis for the same athletes is
also included in Section~\ref{sec:Methods}.  Section~\ref{sec:Results}
section provides the probabilities from the methods section and the paper ends 
by questioning the reaction time barrier.  Lastly, section~\ref{sec:Discussion}
discusses the impact and limitations of the paper.

\section{Data} \label{sec:Data}

There are two types of data that is explored throughout this paper: data from
1999-2022 for every world championship, and data from athletes who competed
at both the 2019 and 2022 World Championships and 2022 national competitions
and the 2022 World Championship.  The first data set is used for the GLMM portion
of the paper, while the second data set is used for the rank based comparisons.

\subsection{World Championship}

\begin{figure}[tbp]
  \centering
  \includegraphics{Finals_Pooled_Boxplot}
  \caption{The reaction times from 2019 to 2022 for the men's final (top) and
  the men's semifinals and finals (bottom).}
  \label{fig:Boxplots}
\end{figure}

Data was taken from the World Athletics and covers the men's 110 meter hurdles 
and the women's 100 meter hurdles from 1999 to 2022.  The variable of prediction 
interest is reaction time, and the predicting variables are year, stage of 
competition (heats, semifinal, final), total time, and gender.  
The data can be found in the appendix~\ref{sec:Appendix}. Factors such as name, 
country, and identification number were not considered important for the 
purposes of this research besides the times for Devon Allen and the other 
United States athletes mentioned above.  The variable ``Stage"
refers to whether the observation occurred during the ``heats" (Preliminary round) which
is denoted by ``H" in the data, ``S" for ``semifinals" (Of which there are either 2 or 3 
heats each year), and ``F" for finals. The variable ``Gender" has levels ``M" and ``F" 
which stand for ``male" and ``female" respectively.  The variable ``Batch" takes a 
numerical value and every heat of every round from 1999 to 2022 was assigned it's
own number.  This was added because of suspected issues that reaction time may be
correlated with other runners reacting quickly (one runner reacting quickly may
result in other runners reacting quickly).  There is no statistical significance to
how the batches were labeled, they were ordered chronologically by meet but
reverse chronologically by year (Batch 1 is a preliminary Heat of 2022 and Batch
111 corresponds to the 1999 finals). Included are
box plots of the reaction times of the men's finals and the men's semifinals and
finals pooled together.  We will refer to the latter as the ``pooled data"
throughout the paper.  The heats or preliminary rounds were not studied because
athletes react faster as the competition progresses.  Athletes react faster in
the finals than the semifinals which are faster than the heats 
\citep{zhang2021correlation}.  If we want to determine if Allen's reaction time
is reasonable, then we thus need to look at the men's final data, with the men's
pooled data being included in order to increase the sample size.  In 2022, there
were only five data points due to two disqualifications and one athlete who 
qualified but did not compete.
The data that looked at in the paper is best visualized using a box plot as show 
in Figure~\ref{fig:Boxplots}.  It is clear that the 2022 box plot for both the
men's finals and semifinals data is lower (indicating a faster reaction time)
than in previous years.  The median for the men's pooled data in 2022 was 0.129
and the median for the men's final data in 2022 was 0.126.  For comparison,
Brosnan found reaction times of 0.156 seconds for men when looking at data from
1999 to 2014 \citep{brosnan2017effects}.

Something that is extremely important to note in the data is the inconsistency 
of World Athletics in terms of how they penalize false starts.  From 2007--2009,
the IAAF (the former name for World Athletics, and the governing body for the 
World Track and Field Champions) instituted a rule change that allowed one 
warning false start before a sprinter was disqualified \citep{iaaf2009falsestart}  This 
essentially gave sprinters a second chance and reduced the penalty for a false 
start.  There were 18 male and 7 female false starts at the 2007 World 
Championships, 18 male and 7 female starts at the 2009 World Championships. 
Starting in 2011, the rule was scrapped, and the old policy which automatically 
disqualifies runners who false start was reinstituted.  The effect? At the 2011 
World Championship there were 6 male disqualifications and 4 female 
disqualifications \citep{iaaf2009falsestart}. By returning to the harsher policy and 
cracking down on false starts, World Athletics had reduced the number of 
false starts by two thirds. One study that looked at different false start
rules that IAAF has imposed from 1997 to 2009 and found that the reaction 
times when there was less leniency improved by a statistically significant amount
\citep{haugen2013effect}. This was desirable for World Athletics as false starts can 
make already long track meets more tedious, both for athletes and viewers watching the 
television broadcast.  Since 2011, IAAF has instituted the one false start rule
\citep{iaaf2009falsestart}.  However, we chose not to include women's data in this
paper due to the numerous papers citing difference in reaction times in men and
women's reaction times: \citep{lipps2011implications, babicc2009reaction,
panoutsakopoulos2020gender}

\subsection{Beyond World Championship}
This section was motivated from a qualitative analysis that started by looking
at how United States athletes reacted at the 2022 United States Track and Field
Championships. From June 23-26 2022, USA held its Track
and Field Championship to decide who to send to the World Championships at Hayward 
Field, the exact same venue as used for the World Track and Field Championship a month 
later.  Thus, a baseline is able to be established for the four USA athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championship races compared to the USA Track and Field 
Championship races. Devon Allen raced three times at the USA Track and
Field Championship and three times at the World Track and Field  Championship. 
His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teammates.  This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
\jy{cross-reference to the table by its label}
too quickly.  Table~\ref{fig:USAvsWorld} below highlights the difference in reaction times for
Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. ``USA" denotes
the USA Track and Field Championships, ``World" denotes the World Track and Field
Championships,   All numbers listed in the table are the reaction
time for the respective athletes in seconds.

\begin{table}
\begin{center}
  \begin{tabular}{c c c c c c c} 
   \toprule
   Athlete & USA H & USA S & USA F & World H & World S & World F \\ [0.5ex] 
   \midrule
   Devon Allen & 0.201 & 0.153 & 0.160 & 0.123 & 0.101 & 0.099 \\ 
   Trey Cunningham & 0.186 & 0.185 & 0.182 & 0.115 & 0.120 & 0.109 \\
   Grant Holloway & 0.192 & 0.190 & N/A & 0.147 & 0.128 & 0.124 \\
   Daniel Roberts & 0.181 & 0.200 & 0.183 & 0.179 & N/A & N/A \\ [0.5ex]
   \bottomrule
  \end{tabular}
  \label{fig:USAvsWorld}
  \caption{Rection times for four United States Athletes at the 2022 USA Track
  and Field World Championships and the 2022 World Track and Field Championship.
  ``N/A" was used for any athlete that did not compete and thus did not register 
  a reaction time.}
  \end{center}
\end{table}

@@

\section{Methods} \label{sec:Methods}

The World Championship data and the data beyond the World Champisonship were
analysed with GLMM and rank-based approaches, respectively.

\subsection{GLMM}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championship, the final model that we chose was a gamma
mixed-effect model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. A random effect for the venue or year was also found to be necessary.
Let $Y_{ij}$ be the reaction time of observation~$j$ in year~$i$.
Conditional on a venue effect $z_i$ of year~i,  $Y_{ij}$ is modeled by 
gamma distribution. In the classic generlized linear model (GLM) sense,
$Y_{ij}$ has conditional mean
\[
g\{E(Y_{ij} | z_i)\} = \alpha + z_i,
\]
and dispersion parameter $\phi$, where $g$ is a known link function and
$z_i$ is a normally distributed random effect with mean zero and
variance~$\sigma^2$. The link function $g$ can be the logarithm function to
ensure the positivity of the conditional mean.
The gamma mixed-effect model can be fit with the \texttt{glmer()} function in R
package \texttt{lme4} \citep{lme4}.

To better understand this model, we can identify the gamma model in terms of the
commonly used shape and scale parameters. For a gamma distribution with
shape~$a$ and scale~$b$, we have mean $\mu = ab$ and variance $v = ab^2$. The
variance as a function of the mean $\mu$ in the GLM sense is
$v(\mu) = \mu^2 / a$, which implies that the dispersion parameter
$\phi = 1 / a$. Therefore, for the gamma mixed-effect model, the conditional
gamma distribution of the reaction time $Y_{ij}$ given the venue effect $z_i$
has shape $a = 1 / \phi$ and scale $b = g^{-1}(\alpha + z_i) \phi$. The marginal
distribution of $Y_{ij}$ is a scale-mixture of gamma distributions, which can be
easily simulated from once the parameters are estimated. A large number of
random numbers generated from the fitted mixture distribution can be used to
approimate the probability of observing a reaction time faster than any given
threshold.  We want the probability of a reaction time being less than 0.1
seconds in order to gauge if that is a reasonable disqualification barrier.

As for the random effect, it was possible to include as well a race effect.
From our exploration, however, the race effect is very
difficult to explain or interpret as it means that some heats of races were either
faster or slower and that everyone in the race reacted similarly.  Although
it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other.  It is much easier however to explain the venue effect.  The technology
that World Athletics has used has differed since 1999 and theoretically the
reaction time data from 2022 should be the best as the timing system should be
state of the art.  Any other weather or climate related factors such as humidity,
precipitation, elevation, etc. would all be negated when looking at the race
effect, but are all factors that could potentially impact the venue effect.

\subsection{Rank-based Comparison}

>>>>

\title{Was Devon Allen Unjustly Disqualified at the 2022 World Track and Field Championships?}

@@

%Abstract should be 200 words
%Devon Allen was disqualified in 2022... discussions about dq, talk methods and 
%models, venue effect, and conclusion
\begin{abstract}
Devon Allen was disqualified in the Men's 110 meter hurdle final of the 2022
Track and Field World Championships after registering a reaction time of 0.099 
seconds, 0.001 seconds faster than what is allowed. Following the games, 
bloggers on the running website LetsRun scrutinized the results %and performed 
%non-statistical analysis of the data. They 
and found that the reaction time data from the 2022 World Track Championships 
seemed to be generally faster compared to the other datasets they looked at, but 
they did not perform any formal %meaningful 
statistical analysis of the data. This paper questions the reaction time 
disqualification barrier, which is currently 0.1 seconds, to determine whether 
this is a reasonable threshold. We employ a generalized linear mixed model 
(GLMM) with a random %effect term which will be referred to as the 
venue effect in order to %try and 
model the data. Additionally, we employ a signed-rank test for clustered data to 
compare reaction times for the same athletes at different competitions. This 
matter needs to be addressed because %Devon Allen's 
disqualification based on allowable reaction time 
%could be a 
%problem that could 
will likely continue to be an issue in future world championships.

\jy{Avoid references in abstract.}

\jy{This is only the background. An abstract also needs to summarize the 
contributions.}
%\bigskip
\eds{removed Devon Allen from keywords because his name is already in title
and replaced with others}
\noindent\sc{Keywords}:
False Start, Hurdles, Reaction Time, Seiko, 2022 World Track and Field Championships

@@

\jy{The bib file needs quality control. The bib keys need to have a consistent
  style. I suggest using the keys from google scholar.}

\jy{Practice my writing tips in the stat-wriging notes: chapter 2 on
  latex. E.g., no vertical/horizontal lines in professional tables. Use the
  tables in my template.}

@@

\jy{Not a good opening. Could start by telling the story, recapping the major
  events and discussions on the internet. Then dig into the relevant
  literature. Finally outline the contributions of this work.
}

\jy{use the following structure and flow:
  1) A paragraph to tell the story;
  2) A paragraph on informal statistical analyses on the internet/social media;
  3) A paragraph on scholarly articles on reaction time in the literature;
  4) A paragraph on what's new: methods; results.
  5) A paragraph of roadmap.}

%1)
After coming in third in the 110 meter hurdles at the United States Track and 
Field Championships in Eugene, Oregon at the end of June, Devon Allen was 
expected to 
be a contender for the 2022 World Track and Field Championships, which were also 
scheduled in Eugene 
two months later. At the World Championships, Allen made it through the 
heats and semifinals to reach the finals while posting reaction times of 0.123
in the heats \eds{how many heats are there? Likely more than one, so I think you
should include all of them, or an average} and 0.101 in the semifinal.  
Allen, who attended the University
of Oregon (the location of US World Championship \eds{US World?} and the World 
Championship) was
considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
in early June; only 0.04 seconds away from the world record \citep{Preview}.
Thus his disqualification by reacting in 0.099 seconds instead of 0.01 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their Youtube
channel that showed the disqualification and can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers 
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting another two years until the next World Track and Field Championship.

%2)
Outside the blog posts from LetsRun.com, there have not been any published
papers specifically regarding Devon Allen's disqualification.  This is not
particularly surprising given that his disqualification occurred only last August.
LetsRun is a website that is part message board and part news. The message board,
which functions similarly to Reddit, is very active during important running
events, including during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as ``Was Devon Allen
Screwed? There's At Least a 99\% Chance That He Was" and ``The Data Keeps Pouring 
In and
It Continues to Look Bad For World Athletics and Great For Devon Allen".  Robert
Johnson, who created and runs the LetsRun website argued that there was an error
with the timing equipment at the 2022 Track and Field World Championship and
cited graphs, simple statistics such as median reaction time, and compared the 
reaction times at the 2022 Track and Field World Championship. 
\eds{compared the reaction times at the 2022 Track and Field World Championship 
to what? Also restate here that no formal statistical analyses have been done.}

%3)
The International Association of Athletics Federations (also referred to
as IAAF or World Athletics), which is the governing board for international running
competitions, commissioned a study in 2009 regarding the reaction time 
barrier and whether the threshold should be lowered. 
\eds{verify previous change is correct} 
The paper analyzed sprinters reaction times and the biological processes 
that went into reacting to the starting gun and concluded 
that athletes can react in as low as 0.08 seconds.  
\eds{what paper are you referring to in the previous sentence?  Please cite.} 
One study by Pain looked at
reaction times for nine male athletes and analyzed their reaction times.  The
fastest athlete of the nine, had on average (even with their two fastest times 
removed) a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. 
\eds{It was recommended (?sentence was previously missing the opening words)}
to IAAF that the disqualification barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body movement.
This could provide a more accurate representation of reaction time, and thus the
possibility of a false disqualification would be lowered \citep{komi2009iaaf}.  
A follow-up study from 2012 by Pilianidis, Mantzouranis, and Kasabalis found
significant differences between reaction times at several world championships
between 1997 and 2009 for the men's 110 meter hurdles.  Another important
conclusion was that reaction times did not decrease over the twelve years
covered by the research in the study \citep{pilianidis2012start}.  This will
be important when considering the results in section \ref{sec:Results}.

Despite the study, IAAF has not changed the reaction barrier since it was
implemented at the 1999 World Track and Field Championships.  It is confusing
as to why IAAF would commission a study only to disregard the results and not
make the necessary changes.  It is unknown how many athletes since 2009 could
have potentially benefited from this, but it is probable that there are others
besides Allen who were disqualified.
%Reaction time barrier has been examined before: cite reviews
%Then go into specifics and talk about people who wanted to change barrier
%General reviews are important
%People have studied the biological processes, cite a few references
%How barrier is picked by regulator
%Read abstract and into and summarize into about one sentence what each paper contributes
%1 and 2 can be merged but 3 need to stand alone

4)
One of the first objectives of the paper is to try to model the data, which is
the reaction times of every non-disqualification since the 1999 World Championships. 
When determining the most effective model to use, several variations were
explored across the 4 different data sets: a linear model without a venue effect,
linear model with venue effect, a gamma model without a venue effect, a gamma model
with a venue effect, a gamma model without a venue effect and link set to log instead
of the default which is Gaussian.  The R package lme4 was very helpful with this
as it supported the Generalized Linear Mixed Model (GLMM) which allowed 
linear models with various distributions to be modeled\citep{Rpkg:lme4}.  The
model we use is a liner model with a fixed effect and random effect and follows
the gamma distribution. The fixed effect comes in the form of an intercept, and the
random effect will be called the venue effect and is reflective of how significant
the changes in the reaction time is for each venue, i.e. the venue effect is 
responsible for the variation between years.

Another objective of this paper is to compare data for athletes who have competed
at additional competitions to see how their reaction times compare from previous
competitions to this one. First we can look at the reaction times of athletes who competed at the
World Track and Field Championship in both 2019 and 2022 and compare the times,
and secondly we can look at how athletes performed compared to how they did at
their national/qualifiers meet.  We would expect that athletes of this caliber,
who are world-class athletes, to not improve significantly from one year to the
next.  In order to reduce any chances of athletes improving reaction time, we
can also look at data from qualifying meets.  Almost every athlete had to qualify 
by proving they are elite at the national level, usually by coming in at least
fourth. Thus, data was compiled on athletes who competed
at the 2022 World Track and Field Championships on what these athletes also ran 
at their country's national championship. If there was any argument that athletes
have improved their reaction time from 2019 to 2022, it should now be considered
nearly null, as the national competitions took place between June and July 2022 
and the World Championships were in August.  Thus, it should be expected that 
elite athletes such as these will not get reasonably faster in such a short time
span. The equipment used to measure reaction time cannot be guaranteed to be
the same from national to international competition, but the time span between
meets has been reduced considerably.
%Simplify contribution to one paragraph: collect data and formal statistical
%analysis such as model used
%Move a lot of this to methods section
5)
The rest of this paper is as follows: Section~\ref{sec:Data} describes how data was
collected and begins to detail the GLMM model which is further developed in the
Section~\ref{sec:Methods}.  The signed-rank test analysis for the same athletes is
also included in the Section~\ref{sec:Methods}.  The Section~\ref{sec:Results}
section provides the probabilities from the methods section and the paper ends 
by questioning the reaction time barrier.

%Follow Anthony's work, explicit references as to what each section does

%Seiko section should go in the Results section.  From results it seems reasonable
%question the data and talk about Seiko

\section{Data} \label{sec:data}

\jy{Describe you data types; give details on sources; introduce notations for
  both types of data
  present exploratory plots.}

  \begin{equation}
    Y_{ij} = \beta_0 + \beta_1 X_{ij} + \gamma_i + \epsilon_{ij}
    \end{equation}

    where $Y_{ij}$ is the response variable for observation $j$ in year $i$, 
    $X_{ij}$ is the predictor variable for observation $j$ in year $i$, 
    $\beta_0$ is the intercept, $\beta_1$ is the coefficient for $X_{ij}$, 
    $\gamma_i$ is the random effect term for year $i$, and $\epsilon_{ij}$ 
    is the error term for observation $j$ in year $i$.

  %yij denotes ith years and j is observation in year
  %yij will have clustered i and we use random effect to account for this
  %When model is presented talk about how I modeled yij
  %Zij where player is i and cluster, group is year and j
  %Test hypothesis that reaction times are the same vs that they are not
Something that is extremely important to note in the Data is the inconsistency of World Athletics 
in terms of how they penalize false starts.  From 2007-2009, the IAAF (the former name for World 
Athletics, and the governing body for the World Track and Field Champions) instituted a rule change
that allowed one warning false start before a sprinter was disqualified \citep{False-Start}  This 
essentially gave sprinters a second chance and reduced the penalty for a false start.  There were 
18 male and 7 female false starts at the 2007 World Championships, 18 male and 7 female starts at 
the 2009 World Championships. Starting in 2011, the rule was scrapped, and the old policy which 
automatically disqualifies runners who false start was reinstituted.  The effect? At the 2011 World
Championship there were 6 male disqualifications and 4 female disqualifications \citep{False-Start}. 
By returning to the harsher policy and cracking down on false starts, World Athletics had reduced the
number of false starts by two thirds. This was desirable for World Athletics as false starts can 
make already long track meets more tedious, both for athletes and viewers watching the 
television broadcast.  Since 2011, IAAF has instituted the one false start rule
\citep{Pilianidis}.

Data was copied from the World Athletics website and pasted into an Excel
spreadsheet. The data covers the men's 110 meter hurdles and the women's 100
meter hurdles from 1999 to 2022.  The variable of prediction interest is reaction
time, and the predicting variables are year, stage of competition (heats, 
semifinal, final), total time, and gender.  The data can be found in the appendix:.  
Factors such as name, country, and identification number were
not considered important for the purposes of this research besides the times for
Devon Allen and the other United States athletes mentioned above.  The variable "Stage"
refers to whether the observation occurred during the "heats" (Preliminary round) which
is denoted by "H" in the data, "S" for "semifinals" (Of which there are either 2 or 3 
heats each year), and "F" for finals. The variable "Gender" has levels "M" and "F" 
which stand for "male" and "female" respectively.  The variable "Batch" takes a 
numerical value and every heat of every round from 1999 to 2022 was assigned it's
own number.  This was added because of suspected issues that reaction time may be
correlated with other runners reacting quickly (one runner reacting quickly may
result in other runners reacting quickly).  There is no statistical signficance to
how the batches were labeled, they were ordered chronologically by meet but
reverse chronologically by year (Batch 1 is a preliminary Heat of 2022 and Batch
111 corresponds to the 1999 finals). The data that looked at in the paper
is best visualized using a box plot. Figure~\ref{fig:DataBoxPlot}.  Included are
box plots of the reaction times of the men's semi finals and finals pooled
together and just the finals.
%Fix quotations
%Dont need the titles will add captions, use width =5, height = 3 in aspect ratio
\begin{figure}[h]
  \centering 
  \includegraphics[width=\textwidth,height=4in]{DataBoxPlot}
  \caption{Boxplot of semifinals and finals data}\label{fig:DataBoxPlot}
\end{figure}
\jy{The plot was stretched. Don't. Get their sizes right before generating them.}

\jy{Remove title from the figures. Make the figure captions self-contained.}
\begin{figure}[h]
  \centering 
  \includegraphics{BoxplotFinals}
  \label{fig:BoxplotFinals}
\end{figure}

This section was motivated from a qualitative analysis that started by looking
at how United States athletes reacted at the 2022 United States Track and Field
Championships. From June 23-26 2022, USA held its Track
and Field Championship to decide who to send to the World Championships at Hayward 
Field, the exact same venue as used for the World Track and Field Championship a month 
later.  Thus, a baseline is able to be established for the four USA athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all of the World 
Track and Field Championship races compared to the USA Track and Field 
Championship races. Devon Allen raced three times at the USA Track and
Field Championship and three times at the World Track and Field  Championship. 
His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teamates.  This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  This table below highlights the difference in reaction times for
Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. "USA" denotes
the USA Track and Field Championships, "World" denotes the World Track and Field
Championships, "N/A" was used for any athlete that did not compete and thus
did not register a reaction time.  All numbers listed in the table are the reaction
time for the respective athletes in seconds.

\begin{center}
  \begin{tabular}{|c c c c c c c|} 
   \hline
   Athlete & USA H & USA S & USA F & World H & World S & World F \\ [0.5ex] 
   \hline\hline
   Devon Allen & 0.201 & 0.153 & 0.160 & 0.123 & 0.101 & 0.099 \\ 
   \hline
   Trey Cunningham & 0.186 & 0.185 & 0.182 & 0.115 & 0.120 & 0.109 \\
   \hline
   Grant Holloway & 0.192 & 0.190 & N/A & 0.147 & 0.128 & 0.124 \\
   \hline
   Daniel Roberts & 0.181 & 0.200 & 0.183 & 0.179 & N/A & N/A \\ [0.5ex]
   \hline
  \end{tabular}
  \end{center}

@@

\section{Methods} \label{sec:meth}

\jy{set up the models/methods using the notations from last section; cite the
  right references.}

\subsection{GLMM}
%Include latex math symbols for the equations
A gamma mixed linear effects model was chosen as the best model to model the data.
Many models were tried and compared to determine what the strongest model.  For
a fixed data set such as the men's pooled data including 2022, the models with
the venue effect included performed better than the models without the venue
effect for each set of comparable models; as indicated by the lower AIC and
higher log likelihood.  Thus, the venue effect is significant and should be
included in the model.  It can also be established that a gamma distribution
better models the data than linear data from comparing the linear model with
venue effect and them gamma model with a venue effect.  The final decision that
was made in model selection was to include link = log in the model.  Despite
doing very little to alter the AIC or log likelihood, the log scale was included
in order to try and better deal with high outliers. %Include Residuals plot?

%Throughout exploratory analysis, gamma provides better AIC
%Spend paragraph on after fitting the model how to figure out 0.1 p value

Once the Gamma mixed effects model was chosen, it needed to be decided which combination
of effects to use.  It was possible to include a venue effect, a batch effect, or a year
and batch effect.  Density plots were created that simulated the probability of a mean
density less than 0.1, which is the current threshold for a reaction time
disqualification.  The density plots are shown below.  The batch effect is very
difficult to explain or interpret as it means that some heats of races were either
faster or slower and that everyone in the batch reacted similarly.  Although
it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other.  It is much easier however to explain the venue effect.  The technology
that World Athletics has used has differed since 1999 and theoretically the
reaction time data from 2022 should be the best as the timing system should be
state of the art.  Any other weather or climate related factors such as humidity,
precipitation, elevation, etc. would all be negated when looking at the batch
effect, but are all factors that could potentially impact the venue effect.

%T, generate from fitted model explain what
%The R code does.  Simulate from fitted model. Use distribution of model to 
%Figure out the p value of reaction time less than 0.1
To try and approximate the distribution of the reaction times, R code was written
to try and estimate the probabilities of observing a reaction time as extreme as
0.1 seconds.  The code re-paramertizes the gamma distribution to be in terms of
$\mu$ and shape rather than by $\alpha$ and $\beta$.

At a size of n=12,000,000, the calculated probability of a reaction
time being below 0.1 seconds was as follows for the various different data sets:

\begin{center}
  \begin{tabular}{|c c|} 
   \hline
   Data Set & Probability RxnTime < 0.1 \\ 
   \hline\hline
   all finals gamma log & 0.000864 \\ 
   \hline
   old finals gamma log & 0.000429 \\
   \hline
   all pooled gamma log & 0.006847 \\
   \hline
   old pooled gamma log & 0.005749 \\
  \end{tabular}
  \end{center}

\subsection{Rank-based comparison for Clustered}

%Citations needed
%Include math formulas and the test statistic
%Equation of test statistic is from papers cited by clusrank. Asymptootic dist
%Was found by __ referecne and the implementation is from the clusrank R package

The test statistic of the DS method of the Wilcoxon rank-sum statistic is:
$Z = \frac{S - E(S)}{\sqrt{\hat{\text{Var}}(S)}}$ where 
$W^* = \frac{1}{N} + \sum_{i=1}^{N} \delta_{i}^{*} R_{i}^{*}$, $R_{i}^{*}$ is
the rank of $X_{i}^{*}$.
The asymptomatic distribution of the Wilcoxon rank-sum statistic was found by
\citet{datta2005rank} and the R package "clusrank" allows for its implementation.
\citep{jiang2017wilcoxon}.
